\vspace*{-8mm}\textbf{\textsc{Novelty Detection for Semantic Place Categorization}}

### 2011-06-23 thesis/novelty-intro.tex
Part 1 of the corrections on chapter 4

The chapter starts by providing the reader with a brief overview on novelty detection
techniques and related works. Then it shows that an optimal novelty detection system
can be implemented by thresholding on a order-relation defined over the inputs.
And that an equivalent order-relation can be imposed by a ratio between
conditional and unconditional probabilities.
Some discussion on how to interpret the meaning of those factors is also
presented.

At last a practical example using semantic data and probabilistic graphical
models is presented: it shows how to use the introduced ratio to obtain a
novelty detection system and analyses the performance impact by increasing
the amount of sensed data and by using an approximation on the unconditional
probability.

\section{Novelty Detection Review and Related Work}
Novelty detection, also known as outlier or anomaly detection, is a
classification problem related to identification of new or unknown data
patterns that the system is not aware of~\cite{markou2003novelty}.

@@

Any normal classification application is also a good candidate for extending
with novelty detection, as the results for samples the system has not been
trained on can be unreliable~\cite{devarakota2008reliability}.
e.g.\ applied to digit-recognition~\cite{tax1998outlier}
or to detection of novel inputs on a neural network~\cite{bishop1994novelty}.

It is in the nature of unknown environments and in the infeasibility of
the system to acquire examples of all possible classes where the complexity of
novelty detection lies: it is only possible to obtain samples representing positives
examples of known cases and the lack of negative examples renders normal
classification methods unusable.
In order to become practical in real world applications, novelty detection
methods have to overcome a series of obstacles:
be able to generalize while still detecting novelty,
be resistant to noisy features,
ability to scale in feature dimension,
deal with multiple classes and performing detection efficiently as many
autonomous systems require real-time or close to real-time performance.

@@

A slightly different approach is applied in case of by one-class \gls{SVM} approaches that
try to distinguish novelty by separating the training set from all the other
points in the input space. They try to achieve that by enclosing the training
set by some structure (e.g.\ an hyper-sphere)~\cite{bennett2000support}.
Those approaches have been made in line with Vapnik principle of not solving
something hard. As although having access to a perfect probability distribution
of the input would solve the problem, creating such a function is harder than
simply creating a boundary between known data and novel data
\cite{scholkopf2000support}.

@@

\cite{ranganathan2010pliss} presents a system able to perform place recognition
and classification from visual clues. It is able to perform segmentation by
exploiting time-coherency on video information. The approach is particularly
interesting by its ability of keeping a fully probabilistic distribution of
the place classification and segmentation. This is, the system never performs
a deterministic decision that impacts any future result, allowing it to adjust
the segmentation and place classification as more data becomes available.
The system is also able to detect novel instances and methods how to adapt it
to run on a constant amount of memory and computation are presented.

\cite{boutell2006factor} presents a method to perform scene-classification
from low-level region detectors using probabilistic graphical models.
The information obtained by the region detectors is exploited together with the
spatial relations between on the higher level scene-classification task.
A presented scheme using only pairwise relations between regions is show to
have better performance than the correct approach of modelling all regions connectivity
with a large single-factor as the pairwise relations can be
approximated with modest amounts of training data, where the single-factor
demands an impractical amount of training data.
No approach is made on it at novelty detection either of regions or scene
categories.

\section{Novelty Detection by Thresholding}
\label{sec:threshold}
The objective of a novelty detection system is to classify a given sample $x$ as
either \emph{known}: $x$ being generated by a known class to the system, or \emph{novel}:
$x$ generated by a class unknown for the system.
Based on the ground\hyp{}truth of a sample four cases are possible:

@@

Due to noisy data, unstable features and lack of information it is impossible to develop
a method able to exactly guess the correct classification of a sample.
By modelling the outcomes with probabilities it becomes possible to handle the uncertainty
associated with each decision.
The notation $P(novel|x)$ will be used to denote the probability that the sample $x$ is in fact a
sample of an unknown class. $P(x)$ will denote the probability that sample $x$ is given to the
system to be classified. $\overline{novel}$ is also defined in such a way that
$P(\overline{novel}|x)$ measures the probability that a decision on classifying $x$ as novel is
wrong.

@@

Note that since probability functions are non-negative, it is impossible to decrease either the
true positive or the false positive probability by using a set $N' \supset N$.
This describes the base of the \emph{error and rejection tradeoff}~\cite{chow1970optimum}, which
states that a system aiming at increasing the true-positive probability will eventually increase its
false-positive error.
The true positive probability can be described as the interest in detecting as much
as possible the novel classes and the false positive probability the interest
in not incurring in too many errors.
By fixating one of those it is possible to define a novelty detection
system that achieves the maximum or minimum of the other.

This way an optimal detector can be formulated by achieving the maximum true-positive
probability without its false-positive probability increase beyond a given limit.
This is equivalent to a \emph{continuous knapsack problem} which allows a greedy
solution by sorting the items with a value per weight function.
In the case of detection system that can be defined as:

\begin{eqnarray}
value(x)  &=& P(\text{true positive}|x) \\
weight(x) &=& P(\text{false positive}|x) \\
cost(x)   &=& value(x)/cost(x) \\
          &=& \frac{P(novel|x)P(x)}{P(\overline{novel}|x)P(x)}
\end{eqnarray}

Therefore a novelty system before classifying a sample $a$ as novel should (greedily)
classify any sample $b$ with a small cost as that would achieve
an higher true positive probability given a fixed false positive one.

@@

Based on this, it can be said that an optimal novelty detection system is
interested in defining an order relation on all the possible inputs equivalent
to the order defined by the function: $P(\overline{novel}|x)$.
And any optimal detector can be described by the largest $P(\overline{novel}|x)$
accepted by it. Which is seen as threshold.

@@

On the previous section it was shown that an optimal novelty detector can be
implemented with a threshold on top of the order-relation defined by
$P(\overline{novel}|x)$ over $x$. Performing some manipulations with
Bayes theorem and assuming a constant $P(\overline{novel})$ a more usable
form can be attained:

@@

Since there is only interest in maintaining the same order-relation as
$P(\overline{novel}|x)$ any constant factor can be dropped.
Leaving a ratio between a \emph{conditional} and
\emph{unconditional probability} suitable for implementing novelty detection
by thresholding.

\subsection{Conditional Probability}
The conditional probability $P(x|\overline{novel})$ describes the distribution
of the samples given that they are generated by a known class.
In case the labelled data available for the agent to learn a concept comes from
the same distribution where the system will run the correct approach is to use it
as prior-information for modelling the conditional probability.

Note that it is important for the labelled data to be a filtered version
of the underlying world distribution (with all classes) that does not contains
any bias.
Otherwise that bias will lead to incorrect modelling the conditional probability
and a subsequent wrong ordering of the sample space.

\subsection{Unconditional Probability}
The unconditional probability $P(x)$ plays an important role on obtaining a
correct order relation for performing novelty detection.
It serves as a normalizing component that allows the system to figure out
whether a given sample conditional probability arises from it belonging to
the known concept or from the likelihood of being sampled.

On lack of any information about the unconditional probability and conforming to
the principle of maximum entropy (\autoref{sec:max-entropy}) a uniform
distribution must be chosen.
Though by using unlabelled data it becomes possible to obtain extra information
and achieve a better approximation.

%\subsection{Novelty Detection on a Variable Set of Features}
Note also that often novelty detection is applied on a fixed set of features
together with an assumption of a uniform unconditional probability.
On those cases $P(x)$ becomes a constant and therefore a novelty threshold
can be directly applied on $P(x|\overline{novel})$ as is the case in \cite{bishop1994novelty}.
But in the case where the set of features $x$ is variable it cannot be
discarded. There $P(x)$ plays a role in levering all the conditional
probabilities on different sets of variables into the same measure units
such that a threshold can be implemented.

@@

>>>>

The chapter starts by providing the reader with a brief overview on novelty detection
techniques and related works. Then it shows that an optimal novelty detection system
can be implemented by thresholding an order relation defined over the inputs.
Moreover, an equivalent order relation can be imposed by a ratio between
conditional and unconditional probabilities.

Finally,  a practical example using semantic data and probabilistic graphical
models is presented: it shows how to use the introduced ratio to obtain a
novelty detection system and analyses the performance impact by increasing
the amount of sensed data and by using an approximation on the unconditional
probability.

\section{Novelty Detection - Review and Related Work}
Novelty detection, also known as outlier or anomaly detection, is a
classification problem related to identification of new or unknown data
patterns that the system is not aware of~\cite{markou2003novelty}.

@@

Any normal classification application is also a good candidate for extending
with novelty detection, as the results for samples on which the system has not been
trained on can be unreliable~\cite{devarakota2008reliability}
e.g. when applied to digit-recognition~\cite{tax1998outlier}
or to detection of novel inputs on a neural network~\cite{bishop1994novelty}.

It is in the nature of unknown environments and in the infeasibility of
the system to acquire examples of all possible classes where the complexity of
novelty detection lies: it is only possible to obtain samples representing positive
examples of known cases and the lack of negative examples renders normal
classification methods unusable.
In order to become practical in real world applications, novelty detection
methods have to overcome a series of obstacles:
be able to generalize while still detecting novelty,
be resistant to noisy features,
scale in feature dimension,
deal with multiple classes and perform detection efficiently as many
autonomous systems require real-time or close to real-time performance.

@@

A slightly different approach is applied in case of by one-class \gls{SVM} approaches that
try to distinguish novelty by separating the training set from all the other
points in the input space. This is achieved by enclosing the training
set with some structure (e.g.\ an hyper-sphere)~\cite{bennett2000support}.
Those approaches have been made in line with Vapnik's principle of never to solve a 
problem which is more general than the one we actually need to solve~\cite{scholkopf2000support}.
Although having access to a perfect probability distribution
of the input would solve the problem, creating such a function is harder than
simply creating a boundary between known data and novel data.

@@

\cite{ranganathan2010pliss} presents a system able to perform place recognition
and classification from visual clues. It is able to perform segmentation by
exploiting time-coherency on video information. The approach is particularly
interesting by its ability of keep a fully probabilistic distribution of
the place classification and segmentation. In other words, the system never performs
a deterministic decision that impacts any future result, allowing it to adjust
the segmentation and place classification as more data becomes available.
The system is also able to detect novel instances and methods how to adapt it
to run on a constant amount of memory and computation are presented.

\cite{boutell2006factor} presents a method to perform scene-classification
from low-level region detectors using probabilistic graphical models.
The information obtained by the region detectors is exploited together with the
spatial relations for the higher level scene-classification task.
A scheme using only pairwise relations between regions is shown to
have better performance than the correct approach of modelling connectivity of all regions 
with a large single factor as the pairwise relations can be
approximated with modest amounts of training data, where the single factor
demands an impractical amount of training data.
No approach is made for novelty detection either of the regions or the scene
categories.

\section{Novelty Detection by Thresholding}
\label{sec:threshold}
The objective of a novelty detection system is to classify a given sample $x$ as
either \emph{known}: $x$ being generated by a class known to the system, or \emph{novel}:
$x$ generated by a class unknown to the system. Based on the ground\hyp{}truth of a sample four cases are possible:

@@

Due to noisy data, unstable features and lack of information, it is impossible to develop
a method able to always exactly guess the correct classification of a sample.
By modelling the outcomes with probabilities, it becomes possible to handle the uncertainty
associated with each decision.
The notation $P(novel|x)$ will be used to denote the probability that the sample $x$ is in fact a
sample of an unknown class. $P(x)$ will denote the probability that sample $x$ is given to the
system to be classified. $\overline{novel}$ is also defined in such a way that
$P(\overline{novel}|x)$ measures the probability that the sample $x$ is generated by a known class.

@@

Note that since probability functions are non-negative, it is impossible to decrease either the
true positive or the false positive probability by using a set $N' \supset N$.
This describes the base of the \emph{error and rejection tradeoff}~\cite{chow1970optimum}, which
states that a system aiming at increasing the true-positive probability will eventually increase its
false-positive error.
The true positive probability is related to the goal of detecting as many novel samples as possible.
At the same time, the false positive probability is related to the goal of not making too many errors.
By fixing one of those, it is possible to define a novelty detection system that achieves the maximum or minimum of the other.

This way an optimal detector can be formulated by achieving the maximum true-positive
probability without its false-positive probability increasing beyond a given limit.
This is equivalent to a \emph{continuous knapsack problem} which allows a greedy
solution by sorting the items with a value per weight function. In the case of 
detection system that can be defined as:

\begin{eqnarray}
value(x)  &=& P(\text{true positive}|x) \\
weight(x) &=& P(\text{false positive}|x) \\
cost(x)   &=& value(x)/weight(x) \\
          &=& \frac{P(novel|x)P(x)}{P(\overline{novel}|x)P(x)}
\end{eqnarray}

Therefore a novelty detection system before classifying a sample $a$ as novel should (greedily)
classify any sample $b$ with a smaller cost as that would achieve a higher true positive probability 
given a fixed false positive one.

@@

Based on this, it can be said that an optimal novelty detection system is
interested in defining an order relation on all the possible inputs equivalent
to the order defined by the function: $P(\overline{novel}|x)$.
And any optimal detector can be described by the largest $P(\overline{novel}|x)$
accepted by it. This can be seen as thresholding.

@@

In the previous section, it was shown that an optimal novelty detector can be
implemented by placing a threshold on the order relation defined by
$P(\overline{novel}|x)$ over $x$. Performing some manipulations with
Bayes theorem and assuming a constant $P(\overline{novel})$, a more usable
form can be attained:

@@

Since the interest is only in maintaining the order relation defined by
$P(\overline{novel}|x)$, any constant factor can be dropped.
Leaving a ratio between a \emph{conditional} and
\emph{unconditional probability} suitable for implementing novelty detection
by thresholding.

\subsection{Conditional Probability}
The conditional probability $P(x|\overline{novel})$ describes the distribution
of the samples given that they are generated by a known class.
In case when the labelled data used to learn a concept come from
the same distribution from the test samples come, the correct approach is to use it
as prior information for modelling the conditional probability.

Note that it is important for the labelled data to be a filtered version
of the underlying world distribution (one modeling all the concepts in the world) 
that does not contain any bias. Otherwise that bias will lead to incorrect modelling of 
the conditional probability and in consequence wrong ordering of the sample space.

\subsection{Unconditional Probability}
The unconditional probability $P(x)$ plays an important role in obtaining a
correct order relation for performing novelty detection.
It serves as a normalizing component that allows the system to decide
whether conditional probability of a given sample arises from it belonging to
the known concept or from the likelihood of being sampled.

On lack of any information about the unconditional probability and conforming to
the principle of maximum entropy (\autoref{sec:max-entropy}) a uniform
distribution must be chosen. However, by using unlabelled data, it becomes possible 
to obtain extra information and achieve a better approximation.

%\subsection{Novelty Detection on a Variable Set of Features}
Note also that often novelty detection is applied to a fixed set of features
together with an assumption of a uniform unconditional probability.
In those cases, $P(x)$ becomes a constant and therefore a novelty threshold
can be directly applied to $P(x|\overline{novel})$ as is the approach presented in \cite{bishop1994novelty}.
However, in case where the set of features $x$ is variable, it cannot be
discarded. There, $P(x)$ plays a role in levering all the conditional
probabilities on different sets of variables into the same units so that a threshold can be obtained.

@@

\begin{sidewaystable}[h]
\begin{center}
\scalebox{0.40}{\input{results/ontology-explain}}
\end{center}
\caption{\label{extra:synthetic-distribution}Distribution used on the synthetic experiment. Each column cell shows $P(feature|class)$}
\end{sidewaystable}
